Welfare Expectations Across U.S. Economic Cycles

### 2024-04-03 paper/paper.qmd
Update first-draft

---
title: "Welfare Expectations Across U.S. Economic Cycles"
subtitle: "Analysis of the the American public's views on government provision of benefits under different economic conditions over the past 50 years"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Welfare_Expectations_and_The_Economy.git. Links marked high-contrast 1:5 ratio green by Professor Alexander in lecture."
date: March 11, 2024
date-format: long
abstract: "A country's economic conditions often influence public opinion about government welfare actions. However, do all members of the public truly align with the general expectation of demanding higher welfare during economic downturns and less so when the economy is thriving? This study examines whether the American public believes government welfare initiatives have been adequate over the past 50 years, especially under varying economic circumstances. We meticulously analyze public sentiment on welfare provision during economic growth, stability, and recession periods, using data from the General Social Survey (GSS) 'Welfare.' Our analysis reveals that the public generally calls for reduced welfare spending during times of economic prosperity, while economic hardship triggers stronger demands for government assistance. Despite this trend, differences in welfare expectations among various demographic groups are evident. For instance, higher-income groups do not advocate for increased welfare support at any time, as this may impact their interests. This finding underscores the need for societies to tailor welfare policies to different groups, informing the development of fairer and more responsive welfare strategies amidst economic fluctuations."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

Leveraging data from the General Social Survey (GSS) that spans decades, this paper examines the complex relationship between economic conditions and public sentiment towards welfare adequacy in the United States. We analyze public opinions during growth, stability, and recession periods to identify overall trends in welfare perceptions and their impact on demographics defined by income and employment status.

By comparing public welfare sentiments during economic downturns with those in stable economic periods, we aim to highlight the nuanced variations in public opinion on welfare provision. We hypothesize that the public expects more benefits during economic lows and fewer benefits during highs. Our preliminary findings reveal a multifaceted landscape of public opinion, with economic downturns significantly influencing the demand for increased government welfare action, albeit with notable variations across different demographic groups.

@@

Structured to offer a comprehensive exploration of these themes, this paper begins with the current introduction that sets the stage for our investigation. The subsequent [Data] section outlines our methodology and data cleaning efforts, preparing the ground for our analysis. In the [Results] section, we present our findings on public welfare perceptions across varying economic conditions. The [Discussion] interprets these findings within the larger context of welfare policy and public expectations, leading to a contemplation of the study's limitations and potential directions for future research. We conclude with a synthesis of our key insights, emphasizing the critical need for nuanced and adaptable welfare policies. An [Appendix] provides further data and survey details, bolstering our analysis.

@@

This article analyzes data from the NORC General Social Survey (GSS) at the University of Chicago, specifically selecting the “welfare” data set from 1972 to 2024. The study enables a comparison of public perceptions of welfare adequacy in different economies. These include 1982, the 2008 global recession and the 2020 recession triggered by the COVID-19 pandemic.

@@

This paper filters out selected variables from selected data variables in GSS and downloads them. The open source statistical programming language R [@citeR] and the libraries `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr clean data], are used in the data process. `tibble` [@tibble], `here` [@here], `kableExtra` [@kableExtra], `janitor` [@janitor] and `knitr` [@knitr].

@@

There are many unclear variable names in the original data set. In order to make the analysis clearer, we renamed the "natfare" variable name to "Response," because this variable mainly indicates the respondents' views on social welfare. Another variable used is "year". The purpose of this variable is to capture the year in which the survey was conducted. Finally, the variable contents of "ballot" and "x5" that are irrelevant to the paper were removed and at the same time, the position structure of the entire dataset was reorganized to better reflect the number of different responses each year..

@@

Time-Series Analysis
In our study, we implemented time-series analysis as a core technique to trace the evolution of public opinions on welfare spending over an extended period. This method allowed us to meticulously map out and evaluate the trajectory of sentiments across decades, identifying any significant trends or sudden shifts in public perception. By analyzing the data as a series of time-ordered points, we were able to discern patterns such as cyclicality or trends corresponding to economic, political, or social events impacting public sentiment towards welfare policies.

Proportion Analysis
To further our understanding of the public's stance on welfare expenditure, we conducted a proportion analysis. This approach involved calculating the percentage of responses falling into predefined categories ('Too Much', 'Too Little', 'About Right') for each survey year. The proportion analysis facilitated a nuanced comparison across different periods, revealing how public sentiment towards welfare spending has fluctuated in response to changes in the socio-economic landscape. It also allowed us to gauge the intensity of public opinion and its alignment or divergence from policy shifts over time.

Ranking and Sorting
Ranking and sorting constituted another pivotal component of our statistical analysis. By categorizing years based on the extremity of sentiments expressed towards welfare spending, we pinpointed periods that stood out due to particularly pronounced public opinions. This method enabled us to isolate and scrutinize years with significant public consensus or discontent regarding welfare policies, offering insights into potential catalysts driving these sentiments and their implications on policy and societal norms.

## Visualization Approach
Line Graphs
For depicting the overarching trends in total survey responses over time, we utilized line graphs. This choice was instrumental in illustrating the engagement level with the subject of welfare spending, capturing fluctuations in public interest and participation in the discourse over the years. The continuous nature of line graphs facilitated a clear visualization of trends, highlighting periods of increased or decreased public engagement, and providing a temporal context for the analysis of sentiment shifts.

Bar Charts
Bar charts played a crucial role in our visualization strategy, especially in representing the proportion of 'Too Much' and 'Too Little' responses. By organizing data into discrete bars, we could effectively showcase the distribution of sentiments for each survey year. The use of color coding and horizontal orientation in certain visualizations was deliberately chosen to enhance the clarity and accessibility of the data, making it easier for readers to grasp the prevalence of specific opinions and their variations over time.

Comparative Visualizations
To draw comparisons between different years and spotlight shifts in public opinion relative to major economic and social events, we designed comparative visualizations. These visual tools were essential in juxtaposing data points from various periods, underscoring the impact of external factors on public sentiment towards welfare spending. Through these visual comparisons, we aimed to provide a more dynamic and contextualized interpretation of the data, enabling readers to connect fluctuations in public opinion with broader socio-economic developments.

# Model Approach

The goal of our modelling strategy is twofold. Firstly,...

@@

@@

@@

>>>>

---
title: "Welfare Expectations Across U.S. Economic Cycles"
subtitle: "Analysis of the the American public's views on government provision of benefits under different economic conditions over the past 50 years"
author: 
  - Tianen (Evan) Hao
thanks: "Code and data are available at: https://github.com/ShadyEvan4830/Welfare_Expectations_and_The_Economy.git. Links marked high-contrast 1:5 ratio green by Professor Alexander in lecture."
date: March 31, 2024
date-format: long
abstract: "A country's economic conditions often influence public perceptions of government welfare practices. However, do all citizens truly meet the general expectation of higher welfare during economic downturns and higher welfare during economic boom? Using data from the General Social Survey (GSS) Welfare, this paper examines whether the American public believes government welfare initiatives over the past 50 years have been adequate, particularly under different economic circumstances. Our analysis shows that there are widespread public calls to reduce welfare spending during good times, while in bad times, this leads to a stronger demand for government assistance. Despite this trend, differences in welfare expectations remain evident between different demographic groups, such as high- and low-income groups. This finding highlights the need for society to develop welfare policies for different groups and provide a basis for creating fairer and more sensitive welfare strategies amid economic fluctuations."
format: pdf
toc: true
number-sections: true
bibliography: bibliography.bib
---

@@

This paper uses nearly 50 years of welfare data from the General Social Survey (GSS) to study the U.S. economic situation and public perceptions of welfare adequacy. We analyze public opinion primarily during periods of growth, stability, and recession to identify overall trends in welfare perceptions and their impact on demographics defined by income and employment status.

We hypothesize that the public expects more benefits when the economy is low and fewer benefits when the economy is growing. Our preliminary findings reveal that public opinion is actually multifaceted. The recession significantly affected demand for increased government welfare action but with significant differences across demographic groups.

@@

The structure of this paper starts with the current [Introduction] to explain the overall logic and content. The subsequent [Data] section outlines our methods and data-cleaning efforts, laying the foundation for our analysis. In the [Results] section, we present findings on public welfare perceptions under different economic conditions. [Discussion] These findings are interpreted within the larger context of welfare policy and public expectations and are expected to trigger reflections on the limitations of the study and potential directions for future research. Finally, we synthesize our main insights in [conclusion] to highlight the urgent need for nuanced and adaptable welfare policies. [Appendix] provides further data and survey details supporting our analysis.

(Evan Hao dreamed about Rohan last night, reminding Evan needed to do a lot cross-reference and graphs/tables later)

@@

This paper analyzes data from the NORC General Social Survey (GSS) at the University of Chicago, specifically selecting the “welfare” data set from 1972 to 2024. The study enables a comparison of public perceptions of welfare adequacy in different economies. These include 1982, the 2008 global recession and the 2020 recession triggered by the COVID-19 pandemic.
\break

@@

This paper filters out selected variables from selected data variables in GSS and downloads them as @tbl-raw-data-overview shown. The open source statistical programming language R [@citeR] and the libraries `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr clean data], are used in the data process. `tibble` [@tibble], `here` [@here], `kableExtra` [@kableExtra], `janitor` [@janitor] and `knitr` [@knitr].

```{r}
#| echo: false
#| tbl-cap: Overview of the Raw Data 
#| label: tbl-raw-data-overview
#| warning: false
#| message: false

# Load packages
library(knitr)
library(kableExtra)

# Read raw data
raw_data <- read.csv("~/Welfare and The Economy/data/raw_data/raw_data.csv")

# Create a table preview with only the first five columns
table_preview <- data.frame(
  Variable = names(raw_data)[1:5],
  NewName = names(raw_data)[1:5],
  Description = c("The year of the survey recorded",
                  "Unique identifier for the respondent",
                  "Respondent's attitude towards national welfare",
                  "Respondent's type of ballot",
                  "Content unclear, column unnamed")[1:5],
  Example = as.character(sapply(raw_data[1:5], function(col) col[1]))
)

# Use kable and kableExtra to create the table
kable(table_preview, "latex", row.names = FALSE, booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = TRUE)  # Only specify LaTeX-compatible styling
```

@@

There are many unclear variable names in the original data set. In order to make the analysis clearer, we renamed the "natfare" variable name to "Response," because this variable mainly indicates the respondents' views on social welfare. Another variable used is "year". The purpose of this variable is to capture the year in which the survey was conducted. Finally, the variable contents of "ballot" and "x5" that are irrelevant to the paper were removed and at the same time, the position structure of the entire dataset was reorganized to better reflect the number of different responses each year as @tbl-cleaned-data-overview shown.

```{r}
#| warning: false
#| tbl-cap: Overview of the Cleaned Data 
#| label: tbl-cleaned-data-overview
#| message: false
#| echo: false

## Cleaned_data Preview
# Load the necessary libraries
library(tidyverse)
library(knitr)
library(kableExtra)

# Load the data
cleaned_data <- read.csv("~/Welfare and The Economy/data/analysis_data/cleaned_data.csv")

# Categorize and summarize the data
categorized_data <- cleaned_data %>%
  count(year, response) %>%
  spread(key = response, value = n, fill = 0) %>%
  mutate(year = as.character(year), # Convert 'year' to character to combine with the 'Total' string later
         Total = rowSums(.[,-1])) # Add a 'Total' column

# Calculate the total row
total_row <- colSums(categorized_data[,-1]) # Sum all columns except 'year'
total_row_df <- as.data.frame(t(total_row))
colnames(total_row_df) <- colnames(categorized_data)[-1] # Ensure column names match
total_row_df$year <- "Total" # Add a 'year' column with the value 'Total'

# Combine the total row with the data frame
categorized_data <- bind_rows(categorized_data, total_row_df)

# Select the top 5 rows for display
categorized_data_top5 <- head(categorized_data, 5)

# Create the table with kable and kableExtra, displaying only the top 5 rows
categorized_data_table <- categorized_data_top5 %>%
  select(year, everything()) %>% # Ensure the 'year' column is first
  kable("latex", row.names = FALSE, booktabs = TRUE) %>%  # Use LaTeX format for PDF
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE) %>%  # Make the header row bold
  column_spec(1, bold = TRUE)  # Make the first column bold

# Print the table
categorized_data_table

```

@@

In our study, we implemented time-series analysis as a core technique to trace the evolution of public opinions on welfare spending over an extended period. This method allowed us to meticulously map out and evaluate the trajectory of sentiments across decades, identifying any significant trends or sudden shifts in public perception. By analyzing the data as a series of time-ordered points, we were able to discern patterns such as cyclicality or trends corresponding to economic, political, or social events impacting public sentiment towards welfare policies.

To further our understanding of the public's stance on welfare expenditure, we conducted a proportion analysis. This approach involved calculating the percentage of responses falling into predefined categories ('Too Much', 'Too Little', 'About Right') for each survey year. The proportion analysis facilitated a nuanced comparison across different periods, revealing how public sentiment towards welfare spending has fluctuated in response to changes in the socio-economic landscape. It also allowed us to gauge the intensity of public opinion and its alignment or divergence from policy shifts over time.

Ranking and sorting constituted another pivotal component of our statistical analysis. By categorizing years based on the extremity of sentiments expressed towards welfare spending, we pinpointed periods that stood out due to particularly pronounced public opinions. This method enabled us to isolate and scrutinize years with significant public consensus or discontent regarding welfare policies, offering insights into potential catalysts driving these sentiments and their implications on policy and societal norms.

## Potential issues with the two-year collection gap in the GSS dataset

When collecting data, we note that the General Social Survey (GSS) collects information every two years. The above initiatives may help the GSS collection process become more accurate and avoid being too time-pressured; however, this interval of data collection may affect our analysis and comparison. For example, if a major policy change or economic event occurs within a year without a survey, we won't be able to immediately see how people react to it. It's like trying to follow a story but missing all the other chapters. Using the welfare analysis targeted at this paper as a reference, if unemployment occurs suddenly in a year for which we have no data, we cannot judge how people feel about their welfare needs during that period. Or, if the economy improves quickly and we don't have data for the current year, our next survey might show that people are okay with receiving less benefits, but we won't know exactly when they start feeling that way. So when we look at our findings, we have to remember that we may not get the full picture.

## Recession Years vs. Stable Economic Years

```{r}
#| warning: false
#| fig-cap: Comparison between 1980 and 1982, 2018 and 2021
#| label: fig-comparison-years
#| message: false
#| echo: false

## Comparison between 1980 and 1982, 2018 and 2021
library(ggplot2)
library(dplyr)
library(readr)

# Load the cleaned data
cleaned_data <- read_csv("~/Welfare and The Economy/data/analysis_data/cleaned_data.csv")

# Filter data for the years and "Too Little" response
comparison_data <- cleaned_data %>%
  filter(year %in% c(1980, 1982, 2018, 2021), response == "Too Little") %>%
  count(year)

# Create the comparison graph
ggplot(comparison_data, aes(x = as.factor(year), y = n, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of 'Too Little' Responses in Selected Years",
       x = "Year",
       y = "Count of 'Too Little' Responses",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "none")  # We do not need a legend since the year is on the x-axis

```

In 1982, the United States grappled with the ramifications of a severe recession that began in the previous year. This downturn was marked by high unemployment, reaching up to 10.8%, and a contraction in the GDP. The elevated bar for 1982 in the graph possibly reflects a heightened public sentiment for more robust welfare support during this trying time.

Moving to 2021, this year was significantly affected by the long-term impacts of the COVID-19 pandemic, which not only caused a public health crisis but also resulted in economic disruptions. Stimulus measures and government support were critical aspects of the economic response. The even higher bar for 2021 than for 1982 may suggest an even greater call from the public for increased welfare, likely influenced by the immediate economic shocks and the pandemic's uncertain long-term outcomes.

Contrastingly, the years 1980 and 2018 are known for more stable economic conditions. In 1980, despite the beginning of a recession, the full impact had not yet been felt, potentially explaining why the bar is lower compared to 1982. The year 2018 is part of a period of sustained growth following the 2008 financial crisis, with lower unemployment rates and a steadily growing GDP. The corresponding bar for 2018 is lower than 2021, indicating fewer responses suggesting that welfare provision was inadequate, aligning with expectations for periods of economic stability.

This visual comparison underscores the interplay between economic conditions and public welfare perceptions. The marked increase in "Too Little" responses during years of economic hardship points to a public demand for greater welfare support when facing economic uncertainty or downturns. Conversely, during stable periods, the pressure on welfare systems appears to be perceived as less intense, reflective in the lower number of "Too Little" responses. The stark contrast between these periods emphasizes the role economic health plays in shaping public opinion on welfare adequacy.

## Concerns about Inconsistent Response Numbers

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

# Calculate the total responses per year
total_responses_per_year <- cleaned_data %>%
  group_by(year) %>%
  summarise(total_responses = n())

## Trends in Public Opinion Over Time in The US
ggplot(total_responses_per_year, aes(x = year, y = total_responses)) +
  geom_line(color = "#2C3E50", size = 1) +
  geom_point(color = "#E74C3C", size = 3, shape = 18) +
  theme_minimal(base_size = 14) +
  labs(title = "Trend of Total Responses per Year",
       x = "Year",
       y = "Total Responses",
       caption = "Annual trend showing the total number of responses") +
  theme(plot.title = element_text(hjust = 0.5))

```

(Evan Hao dreamed about Rohan last night, reminding him that he needed to do a cross-reference) The graph provided displays the annual trend of total responses from the General Social Survey (GSS). Upon examination, there's a noticeable fluctuation in the number of responses over the years. These inconsistencies raise concerns about the stability and reliability of the dataset, which could influence the robustness of any analyses conducted using this data.

The peaks and troughs in the graph could suggest varying levels of engagement or interest in the survey across different years. For example, the significant dip around the early 1990s could indicate external factors that might have affected participation rates, such as social or political events. Conversely, the sharp increase in recent years, such as the noticeable peak in 2020, might reflect heightened public interest in welfare issues, perhaps spurred by the COVID-19 pandemic's economic impact.

For a researcher, these inconsistencies could complicate trend analysis. If the number of responses is not consistent, it's challenging to determine whether changes in the data are due to actual shifts in public opinion or simply the result of varying sample sizes. For instance, a lower number of total responses in a given year could artificially deflate the perceived importance of welfare issues if not properly accounted for in the analysis. Similarly, a sudden increase in responses might overemphasize concerns that are not as pressing across the broader population.

When using this data, it's essential to consider these variations and apply statistical techniques that can mitigate their impact. This might include weighting responses to balance out the differences or conducting sensitivity analyses to ensure that the observed trends are not artifacts of the data collection process.

# Model

In order to better analyze the trend, we analyzed the trend of replying to "Too Little" through the model. . . The structure shows that the trend has been rising in the past 50 years.
The goal of our modelling strategy is twofold. Firstly,...

@@

```{r}
#| warning: false
#| tbl-cap: Linear model showing the trend of Too Little responses
#| label: tbl-linear-model
#| message: false
#| echo: false
#| eval: true

library(tidyverse)
library(rstanarm)
library(modelsummary)

# Load the model (make sure the path is correct and use `here` if needed)
#model_path <- "~/Welfare and The Economy/models/analysis_of_the_trend_model.rds"
model_path <- "~/Welfare and The Economy/models/first_model.rds"
first_model <- readRDS(file = model_path)

too_little_data <- read_csv("~/Welfare and The Economy/data/analysis_data/too_little_data.csv")

# Print out that the model was loaded successfully for debugging purposes
# cat("Model loaded successfully from:", model_path, "\n")

```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: tbl-lalalala
#| tbl-cap: Linear model showing the trend of Too Little responses

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
# Visualizing the trend
ggplot(too_little_data, aes(x = year, y = proportion_too_little)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Trend of 'Too Little' Responses Over Years",
       x = "Year",
       y = "Proportion of 'Too Little' Responses",
       caption = "Linear model showing the trend of 'Too Little' responses") +
  theme_minimal()

```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

# Load cleaned data
cleaned_data <- read.csv("~/Welfare and The Economy/data/analysis_data/cleaned_data.csv")

# Calculate the total number of responses per year
total_responses_per_year <- cleaned_data %>%
  group_by(year) %>%
  summarise(total = n())

# Calculate the number and proportion of "Too Much" responses per year
too_much_proportion <- cleaned_data %>%
  filter(response == "Too Much") %>%
  count(year) %>%
  rename(num_responses = n) %>%
  inner_join(total_responses_per_year, by = "year") %>%
  mutate(proportion = num_responses / total) %>%
  arrange(desc(proportion))
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

## Years Ranked by the Highest Proportion of "Too Little" Responses
library(tidyverse)

# Load cleaned data
cleaned_data <- read.csv("~/Welfare and The Economy/data/analysis_data/cleaned_data.csv")

# Assuming cleaned_data has a 'response' column where 'Too Little' responses are indicated
# Calculate the proportion of 'Too Much' responses for each year
too_much_proportion <- cleaned_data %>%
  filter(response == "Too Little") %>%
  count(year) %>%
  mutate(total_responses = sum(n), proportion = n / total_responses) %>%
  select(year, proportion)
```

@@

# Conclusion

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```
